Emulation of high multiplicity matrix elements for electron-positron and hadron-hadron colliders

### 2022-09-09 intro.tex
qcd, scattering amplitudes, cross-sections

\section{Introduction to quantum chromodynamics}
\begin{itemize}
    \item QCD is the quantum field theory describing the strong
    interaction between quarks and gluons. It is a non-abelian
    gauge theory with gauge group SU(3).
    \item Write down the Lagrangian of QCD and define all the
    terms (field strength tensor, quark fields, group generators,
    structure constants, coupling constants.)
    \item Begin to talk about how Feynman rules can be used
    to represent the interactions that you can derive from
    the Lagrangian.
\end{itemize}
    Quantum chronodynamics (QCD) is the sector of the SM that
    describes the strong interaction. 
    QCD is a non-abelian gauge theory with gauge group
    $\mathrm{SU}(3)_{C}$ where the charge is named colour.
    The gauge and fermion part of the QCD Lagrangian is
    \begin{equation}\label{eqn:L_QCD}
        \mathcal{L}_{\mathrm{QCD}} = -\dfrac{1}{4}F^{a}_{\mu\nu}F^{a, \, \mu\nu}
        + \sum_{f} \bar{\psi}_{i}^{f}(\mathrm{i}\slashed{D}_{ij} - \delta_{ij}m_{f})\psi_{j}^{f} \, ,
    \end{equation}
    where twice repeated indices are summed over.
    The fields $\psi_{i}^{(f)}$ ($\bar{\psi}_{i}^{f}$) are the fermions (antifermions),
    representing quarks (antiquarks) with flavours $f$: 
    up $u$, down $d$, strange $s$, charm $c$, top $t$, and bottom $b$,
    with masses $m_{f}$. They transform under the fundamental
    representation with indices $i, j \in \{1, 2, 3\}$, named colour indices.
    The gauge fields $A^{a}_{\mu}$, corresponding to gluons, 
    appear in the quark covariant derivative
    \begin{equation}\label{eqn:covariant_deriv}
        (D_{\mu})_{ij} = \delta_{ij}\partial_{\mu} - ig_{\mathrm{s}}T^{a}_{ij}A^{a}_{\mu} \, ,
    \end{equation}
    and the gauge field strength tensor
    \begin{equation}\label{eqn:F_munu}
        F^{a}_{\mu\nu} = \partial_{\mu}A^{a}_{\nu} - \partial_{\nu}A^{a}_{\mu} + g_{\mathrm{s}} f^{abc}A^{b}_{\mu}A^{c}_{\nu} \, .
    \end{equation}
    Gauge fields transform under the adjoint representation, meaning
    the adjoint indices $a,b,c \in \{1,...,8\}$. $T^{a}_{ij}$ are the
    group generators in a fundamental representation. In SU(3) it is
    common to write the group generators as
    \begin{equation}\label{eqn:group_generators}
        T^{a}_{ij} = \dfrac{1}{2}\lambda^{a}_{ij}
    \end{equation}
    where $\lambda^{a}_{ij}$ are the Gell-Mann matrices \cite{Gell-Mann:1962yej}.
    The generators of the group have to obey the Lie algebra
    \begin{equation}\label{eqn:lie_algebra}
        [T^{a}, T^{b}] = \mathrm{i}f^{abc}T^{c} \, ,
    \end{equation}
    where $f^{abc}$ are the structure constants of SU(3).
    The gauge coupling of the group $g_{\mathrm{s}}$
    is a dimensional free parameter of the theory, usually
    written as
    \begin{equation}\label{eqn:alpha_s}
        \alpha_{\mathrm{s}} = \dfrac{g_{\mathrm{s}}^{2}}{4\pi} \, ,
    \end{equation}
    for reasons which will become clear when we introduce
    scattering amplitudes.

\subsection{Running coupling}
\begin{itemize}
    \item Strength of the coupling depends on the energy
    that the process occurs at. This is known as the running
    of the coupling.
    \item The value of $\alpha_{s}$ is usually quoted
    at the mass of the Z boson.
    \item In QCD the evolution of $\alpha_{s}$ is governed
    by the renormalisation group equation (Callan-Symanzik
    $\beta$ function). The coefficients of this function
    are known to many loops.
    \item Asymptotic freedom is a property of non-abelian
    gauge groups and this discovery won a Nobel prize.
\end{itemize}
\subsection{Scattering amplitudes}
\begin{itemize}
    \item S-matrix/Transition matrix
    \item Make the link between Feynman diagrams and
    scattering amplitudes/matrix elements. The sum
    of all Feynman diagrams you can draw for a particular
    process is the matrix element.
    \item Explicitly state how the QCD Feynman rules are
    derived and have the corresponding diagrams.
    \item Explain how you can basically join these
    Feynman rules into a diagram which represents a
    convenient way to calculate interactions between
    particles.
    \item Introduce the concept of loop amplitudes here,
    which naturally leads onto divergences.
    \item UV divergences lead to the introduction of
    dimensional regularisation and also to renormalisation
    of fields to make them represent physical quantities.
    The main scheme that people use is the modified minimal
    subtraction scheme.
\end{itemize}
\section{Experiments and Phenomonology}
\subsection{Observables}
\begin{itemize}
    \item Experiments do not measure single particles,
    they measure hadrons which is far from the collision
    of protons.
    \item Factorisation of hard and soft scales leads
    to the concepts of PDFs (DGLAP equation), partonic
    cross-sections (Feynman diagrams), parton showers,
    and hadronisation.
    \item Link back to matrix elements and how they
    are integrated to give you cross-sections which
    can be compared with experiments.
    \item Probably need to talk about infrared safe
    observables and jets.
\end{itemize}
\section{Monte Carlo event generator overview}
\subsection{Event unweighting}
\begin{itemize}
    \item MC generators are the main tools that
    people use to compare with experimental data.
    There are many groups that have their own tool.
    \item The usage of MC tools came about because
    the multi-dimensional phase-space integrals
    are too complex to do analytically.
    \item This doesn't mean that they only do
    the integrals, the event generators basically
    contain all the tools from phase-space generation
    to matrix element evaluation, to parton showers
    to hadronisation.
    \item Have a look at Katherina thesis to see
    how much I can skip because only really want to
    introduce unweighting. Probably along the lines
    of event weight is statistically sampled vs
    what they measure in real life. Therefore we need
    to unweight events.
\end{itemize}
\end{document}

>>>>

\section{Introduction to quantum chromodynamics}
\begin{itemize}
    \item QCD is the quantum field theory describing the strong
    interaction between quarks and gluons. It is a non-abelian
    gauge theory with gauge group SU(3).
    \item Write down the Lagrangian of QCD and define all the
    terms (field strength tensor, quark fields, group generators,
    structure constants, coupling constants.)
    \item Begin to talk about how Feynman rules can be used
    to represent the interactions that you can derive from
    the Lagrangian.
\end{itemize}
    Quantum chronodynamics (QCD) is the sector of the SM that
    describes the strong interaction. 
    QCD is a non-abelian gauge theory with gauge group
    $\mathrm{SU}(3)_{C}$ where the charge is named colour.
    The gauge and fermion part of the QCD Lagrangian is
    \begin{equation}\label{eqn:L_QCD}
        \mathcal{L}_{\mathrm{QCD}} = -\dfrac{1}{4}F^{a}_{\mu\nu}F^{a, \, \mu\nu}
        + \sum_{f} \bar{\psi}_{i}^{f}(\mathrm{i}\slashed{D}_{ij} - \delta_{ij}m_{f})\psi_{j}^{f} \, ,
    \end{equation}
    where twice repeated indices are summed over.
    The fields $\psi_{i}^{f}$ ($\bar{\psi}_{i}^{f}$) are the fermions (antifermions),
    representing quarks (antiquarks) with flavours $f$: 
    up $u$, down $d$, strange $s$, charm $c$, top $t$, and bottom $b$,
    with masses $m_{f}$. They transform under the fundamental
    representation with indices $i, j \in \{1, 2, 3\}$, named colour indices.
    The gauge fields $A^{a}_{\mu}$, corresponding to gluons, 
    appear in the quark covariant derivative
    \begin{equation}\label{eqn:covariant_deriv}
        (D_{\mu})_{ij} = \delta_{ij}\partial_{\mu} - ig_{\mathrm{s}}T^{a}_{ij}A^{a}_{\mu} \, ,
    \end{equation}
    and the gauge field strength tensor
    \begin{equation}\label{eqn:F_munu}
        F^{a}_{\mu\nu} = \partial_{\mu}A^{a}_{\nu} - \partial_{\nu}A^{a}_{\mu} + g_{\mathrm{s}} f^{abc}A^{b}_{\mu}A^{c}_{\nu} \, .
    \end{equation}
    Gauge fields transform under the adjoint representation, meaning
    the adjoint indices $a,b,c \in \{1,...,8\}$. $T^{a}_{ij}$ are the
    group generators in a fundamental representation. In SU(3) it is
    common to write the group generators as
    \begin{equation}\label{eqn:group_generators}
        T^{a}_{ij} = \dfrac{1}{2}\lambda^{a}_{ij}
    \end{equation}
    where $\lambda^{a}_{ij}$ are the Gell-Mann matrices \cite{Gell-Mann:1962yej}.
    The generators of the group have to obey the Lie algebra
    \begin{equation}\label{eqn:lie_algebra}
        [T^{a}, T^{b}] = \mathrm{i}f^{abc}T^{c} \, ,
    \end{equation}
    where $f^{abc}$ are the structure constants of SU(3).
    The gauge coupling of the group $g_{\mathrm{s}}$
    is a dimensional free parameter of the theory, usually
    written as
    \begin{equation}\label{eqn:alpha_s}
        \alpha_{\mathrm{s}} = \dfrac{g_{\mathrm{s}}^{2}}{4\pi} \, ,
    \end{equation}
    for reasons which will become clear when we discuss
    scattering amplitudes.

    To remove unphysical degrees of freedom from the
    theory we need to gauge-fix the theory
    and add a ghost Lagrangian. The gauge-fixing term
    in the $R_{\xi}$ gauges is written as
    \begin{equation}\label{eqn:L_GF}
        \mathcal{L}_{\mathrm{GF}} = -\dfrac{1}{2\xi}(\partial^{\mu}A_{\mu}^{a})^{2} \, ,
    \end{equation}
    where $\xi = 1$ corresponds to the Feynman-'t-Hooft gauge.
    The ghost Lagrangian is most commonly written
    in the Faddeev-Popov procedure as
    \begin{equation}\label{eqn:L_ghost}
        \mathcal{L}_{\mathrm{ghost}} = (\partial_{\mu}\bar{c}^{a})(\delta^{ac}\partial_{\mu} + g_{\mathrm{s}}f^{abc}A^{b}_{\mu})c^{c} \, ,
    \end{equation}
    where $c^{a}$ ($\bar{c}^{a}$) are Faddeev-Popov ghosts (anti-ghosts).
    They are unphysical external states but must be
    included in internal lines for a Lorentz invariant
    perturbative gauge theory.

    Now that we have written down the QCD Lagrangian,
    we can examine the terms to write down how the
    gauge bosons interact with the fermions. A convenient
    way to do this is with Feynman rules, which can be
    pieced together to form Feynman diagrams. We will see
    that Feynman diagrams are a simple, but powerful, tool
    to systematically build all the possible ways a process
    can occur. This will be discussed further in the
    scattering amplitude section.

    Expanding (\ref{eqn:L_QCD}) and extracting the terms
    that mix the gauge and fermions fields, we get an interaction Lagrangian
    \begin{equation}\label{L_int}
        \mathcal{L}_{\mathrm{int}} = g_{\mathrm{s}}A^{a}_{\mu}\bar{\psi}_{i}^{f}\gamma^{\mu}T^{a}_{ij}\psi_{j}^{f}-g_{\mathrm{s}}f^{abc}(\partial_{\mu}A^{a}_{\nu})A^{b,\,\mu}A^{c,\,\nu} -\dfrac{g_{\mathrm{s}}^{2}}{4}f^{abc}A^{b}_{\mu}A^{c}_{\nu}f^{ade}A^{d,\,\mu}A^{e,\,\nu} \, ,
    \end{equation}
    which we can interpret as follows: the first term
    is an interaction between a gluon, a quark, and an
    anti-quark, the second term is a triple gluon
    self-interaction and the third term is a four gluon
    self-interaction. These mixing terms can be recast
    as Feynman rules as interaction vertices.

    To connect vertices we require propagators, which
    we can read off from the kinetic Lagrangian where
    we collect terms from (\ref{eqn:L_QCD}) and (\ref{eqn:L_GF}), 
    \begin{equation}\label{L_kin}
        \mathcal{L}_{\mathrm{kin}} = -\dfrac{1}{4}(\partial_{\mu}A^{a}_{\nu}-\partial_{\nu}A^{a}_{\mu})^{2} - \dfrac{1}{2\xi}(\partial_{\mu}A^{a}_{\mu})^{2} + \bar{\psi}_{i}^{f}(\mathrm{i}\slashed{\partial}-m_{f})\psi_{i}^{f} \, ,
    \end{equation}
    where the first two terms corresponds to a gluon
    propagator and the third term corresponds to
    a quark propagator.

    Note that we have only written down the vertices
    and propagators for the physical gluons and quarks,
    ghosts also have associated Feynman rules but we do
    not collect those here. A full list of Feynman rules
    in the SM can be found at Ref. \cite{Romao:2012pq}.

    Now that we have the Feynman rules for the theory
    we can move on to discuss scattering amplitudes
    and cross-sections.

\subsection{Scattering amplitudes and cross-sections}
    In collider experiments we typically collide
    two beams, consisting of bundles of energetic
    particles, and analyse the resulting products
    of these collisions. To model collisions in
    a collider experiment, we look at the specific
    case of two particles colliding into $n$ particles.
    To mathematically describe these
    scattering processes, scattering amplitudes are used.
    For an initial-state
    $|i\rangle$ and final-state $|f\rangle$
    the scattering amplitude can be written as
    \begin{equation}\label{eqn:scattering}
        \langle f | S | i \rangle \, ,
    \end{equation}
    where the scattering matrix, or $S$-matrix
    can be decomposed into an identity matrix
    and a transfer matrix $\mathcal{T}$,
    \begin{equation}\label{eqn:S_matrix}
        S = \mathbbm{1} + \mathrm{i}\mathcal{T} \, .
    \end{equation}
    The $S$-matrix encodes all the information
    about how the initial-state will evolve
    into the final-state over time. By writing
    the $S$-matrix in this way, all the interactions
    are separated into the transfer matrix $\mathcal{T}$,
    as the identity matrix describes the free theory.
    By imposing a momentum conserving constraint
    on the $S$-matrix, the matrix element,
    $\mathcal{M}$, is defined in the following expression
    \begin{equation}\label{eqn:matrix_element}
        \langle f | \mathcal{T} | i \rangle = (2\pi)^{4} \delta^{4}\left(p_{a} + p_{b} - \sum_{f=1}^{n} p_{f}^{\mu}\right) \mathcal{M} \, .
    \end{equation}
    where we take $p_{a}$ and $p_{b}$ to be
    the momenta of the two colliding
    particles in the initial-state, and $p_{f}$ to be the
    momenta of the $n$-particle final states.
    The transition probability of this scattering process
    is the modulus squared of the scattering amplitude
    \begin{equation}\label{eqn:S_prob}
        P = |\langle f | S | i \rangle|^{2} \propto |\langle f | \mathcal{M} | i \rangle 
    \end{equation}
    where $|\langle f | \mathcal{M} | i \rangle|^{2}  \equiv |\mathcal{M}|^{2}$
    is the matrix element squared.

    Matrix elements can be calculated from
    first principles in QFT using Feynman diagrams.

    Since detectors at collider experiments
    have limited resolution, single scattering
    events are not measured. Instead we consider
    the cross-section, $\sigma$, which is a property
    of the particles being scattered, and is independant
    of the way the experiment is carried out.
    The cross-section directly relates the
    matrix elements which we calculate in our QFT,
    to measurable observables at experiments.
    In practice, it is more useful to consider the
    differential cross-section, $\mathrm{d}\sigma$
    where the cross-section can be differential
    in quantities such as energies and angles.
    This is because the final-state particles
    will be produced with a range of, for example,
    energies and momenta so the cross-section would
    be vanishing for an exact momentum configuration.

    The differential cross-section can be written as
    \begin{equation}\label{eqn:dsigma}
        \mathrm{d}\sigma = \dfrac{1}{\mathcal{F}}|\mathcal{M}_{2 \rightarrow n}|^{2} \mathrm{d}\Phi_{n} \, .
    \end{equation}
    $\mathcal{F}$, known as the flux factor
    is
    \begin{equation}\label{eqn:flux}
        \mathcal{F} = 4\sqrt{(p_{a}p_{b})^{2} - m_{a}^{2}m_{b}^{2}} \xrightarrow[\mathrm{limit}]{\mathrm{massless}} 2(p_{a} + p_{b})^{2} = 2s_{ab} \, .
    \end{equation}

    $\mathrm{d}\Phi_{n}$ is the Lorentz-invariant
    phase-space which contains all the possible
    configurations of the $n$-particle final state.
    Absorbing the momentum conserving $\delta$-function
    from (\ref{eqn:matrix_element}),
    we can write it as
    % \begin{equation}\label{eqn:dlips}
    %     \mathrm{d}\Phi_{n} = (2\pi)^{d}\delta^{(d)}\left(p_{a} + p_{b} - \sum_{f} p_{f}^{\mu}\right) \prod_{f=1}^{n} \dfrac{\mathrm{d}^{d}p_{f}}{(2\pi)^{d-1}}\delta(p_{f}^{2} - m_{f}^{2})\theta(E_{f}) \, ,
    % \end{equation}
    \begin{equation}\label{eqn:dlips_4d}
        \mathrm{d}\Phi_{n} = (2\pi)^{4}\delta^{4}\left(p_{a} + p_{b} - \sum_{f=1}^{n} p_{f}^{\mu}\right) \prod_{f=1}^{n} \dfrac{\mathrm{d}^{4}p_{f}}{(2\pi)^{3}}\delta(p_{f}^{2} - m_{f}^{2})\theta(E_{f}) \, ,
    \end{equation}
    where the second $\delta$-function restricts
    the final state particles to be on mass-shell
    (on-shell). The step function selects only
    the positive energy solution for this constraint.
% \begin{itemize}
%     \item Strength of the coupling depends on the energy
%     that the process occurs at. This is known as the running
%     of the coupling.
%     \item The value of $\alpha_{s}$ is usually quoted
%     at the mass of the Z boson.
%     \item In QCD the evolution of $\alpha_{s}$ is governed
%     by the renormalisation group equation (Callan-Symanzik
%     $\beta$ function). The coefficients of this function
%     are known to many loops.
%     \item Asymptotic freedom is a property of non-abelian
%     gauge groups and this discovery won a Nobel prize.
% \end{itemize}
\section{Experiments and Phenomonology}
\subsection{Observables}
\begin{itemize}
    \item Experiments do not measure single particles,
    they measure hadrons which is far from the collision
    of protons.
    \item Factorisation of hard and soft scales leads
    to the concepts of PDFs (DGLAP equation), partonic
    cross-sections (Feynman diagrams), parton showers,
    and hadronisation.
    \item Link back to matrix elements and how they
    are integrated to give you cross-sections which
    can be compared with experiments.
    \item Probably need to talk about infrared safe
    observables and jets.
\end{itemize}
\section{Monte Carlo event generator overview}
\subsection{Event unweighting}
\begin{itemize}
    \item MC generators are the main tools that
    people use to compare with experimental data.
    There are many groups that have their own tool.
    \item The usage of MC tools came about because
    the multi-dimensional phase-space integrals
    are too complex to do analytically.
    \item This doesn't mean that they only do
    the integrals, the event generators basically
    contain all the tools from phase-space generation
    to matrix element evaluation, to parton showers
    to hadronisation.
    \item Have a look at Katherina thesis to see
    how much I can skip because only really want to
    introduce unweighting. Probably along the lines
    of event weight is statistically sampled vs
    what they measure in real life. Therefore we need
    to unweight events.
\end{itemize}
\end{document}
